The YUIMA Project: a Computational Framework for Simulation and Inference of Stochastic Differential Equations

### 2013-10-10 pkg/yuimadocs/inst/doc/JSS/final-edited.Rnw
print methods
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/yuima@250 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The general idea of the \pkg{yuima} package is to separate into different subclass objects the statistical model, the data and the statistical methods. As will be explained with several examples, the user may give a mathematical description of the statistical model with \code{setModel} which prepares a \code{yuima.model} object by filling the appropriate slots. If the aim is the simulation of the solution of the stochastic  differential equation specified in the \code{yuima.model} object then, using the method \code{simulate}, it is possible to obtain one trajectory of the process. As an output, a \code{yuima} object (by ``\code{yuima} object'' we mean a, possibly incomplete, object of class \pkg{yuima}) is created which contains the original model specified in the \code{yuima.model} object in the slot named \code{model} and two additional slots named \code{data}, for the simulated data, and \code{sampling} which contains the description of the simulation scheme used as well as other informations. The details of \code{simulate} will be explained in Section \ref{sec:simul} along with the use of method \code{setSampling} which allows to specify the type of sampling scheme to be used by the \code{simulate} method.
But \code{yuima} object may contain the slot \code{data} non only as the outcome of \code{simulate} but also because the user decides to analyse its own data. In this case the method \code{setData} is used to transform most types  of \proglang{R} time series object into a a proper \code{yuima.data} object.
When the slots \code{data} and \code{model} are available, many other methods can be used to perform statistical analysis on these SDE's (stochastic differential equation) models. These methods will be discussed in Section \ref{sec5}.
Further, functionals of stochastic differential equations can be defined using the \code{setFunctional} method and evaluated using asymptotic expansion methods as explained in Section \ref{sec4}. The \code{setFunctional} method creates a \code{yuima.functional} object which is included along with a \code{yuima.model}  into a \code{yuima} object in order to be used for the evaluation of its asymptotic expansion.

\subsection{The \code{yuima.model} Class}\label{sec:model}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified.  Here we present  a brief
overview of these models as they will be described in detail in Section \ref{sec3}, but this allows to introduce an overall view of the slots of the \code{yuima.model} class.
In \pkg{yuima} one can describe three main families of stochastic processes at present. These models can be one or multidimensional and eventually described as parametric models. Let $X_0=x_0$ be the initial value of the process, then, the main classes are as follows:
\begin{itemize}
\item diffusion models described as
$$  \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t, \quad X_0=x_0$$
 where $W_t$ is a standard Brownian motion;
\item SDE's driven by fractional Gaussian noise, with $H$ the Hurst parameter, described as
$$ \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t^{H}, \quad X_0=x_0;$$ 
\item diffusion process with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0;
$$
\end{itemize}
 The functions $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ may have a different number of arguments. For example, if the model is homogeneous in time and drift and diffusion coefficients are entirely specified, then we will use the notaion $a(x)$ and $b(x)$ and describe the diffusion model simply as $\displaystyle \de X_t=a(X_t)dt + b(X_t)\de W_t$. And so forth. Detailed hypotheses and regularity conditions on the coefficients $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ for each class of models will be given in the next sections. Nevertheless, it is important to remark that these notations only matter to the mathematical description of the model while each coefficient is passed to \pkg{yuima} methods as \proglang{R} mathematical expressions. This means that, for example, $a(t, X_t, \theta) = t\cdot \sqrt{\theta X_t}$ will be passed as \code{t*sqrt(x*theta)}, therefore, the order of the arguments is not relevant to \proglang{R} as well as its mathematical description as long as it is consistent through each specific section. Furhter, \pkg{yuima} is able to accept any user-specified notation for the state variable $x$ (for $X_t$) and the time variable $t$ and the remaining term of an \proglang{R} expression will be interpreted as parameter as will be explained in Section \ref{sec:diff}.
We are now able to give an overview of the main slots of the most important class of the \pkg{yuima} package.

@@

To see inside its structure, one can use the \proglang{R} command \code{str} on a \pkg{yuima} object.

@@

As seen in the above, the parameter space is accurately described internally in a \code{yuima} object because in inference for stochastic differential equations, estimators of different parameters have different properties. Usually, the rate of convergence for estimators in the diffusion coefficient are similar to the ones in the i.i.d. (independent and idensitcally distributed) sampling while estimators of parameters in the drift coefficient are slower or, in some cases, not even consistent. The \pkg{yuima} always tries to implement the best statistical inference for the given model under the observed sampling scheme.

@@

\subsection{One Dimensional Diffusion Processes}\label{sec:diff}
Assume that we want to describe the following stochastic differential equation
$$\de X_t = -3 X_t \de t + \frac{1}{1+X_t^2}\de W_t.$$
In the above $a(x) = -3 x$ and $b(x) = \frac{1}{1+x^2}$ according to the notaion of previous section and $W_t$ is a standard Wiener process.
This can be described in \pkg{yuima} by specifying the drift and diffusion coefficients as plain
\proglang{R} expressions passed as strings
<<print=FALSE,echo=FALSE,results=hide>>=
library(yuima)
@ 
<<echo=TRUE, print=FALSE,results=hide>>=
mod1 <- setModel(drift = "-3*x", diffusion = "1/(1+x^2)")
@ 
By default, \pkg{yuima} assumes that the state variable is \code{x} and the time variable is \code{t} and the solve variable is again \code{x}. Notice that the left hand-side of the equation is implicit, this is why \code{yuima.model} has the slot \code{solve.variable}.
The user should not be worried about the warning raised by \pkg{yuima} at this stage, as this is just to inform her on the implicit assumption on the solution variable.
At this point, the package fills the proper slots of the \code{yuima} object
<<>>=
str(mod1)
@

>>>>

The general idea of the \pkg{yuima} package is to separate into different subclass objects the statistical model, the data and the statistical methods. As will be explained with several examples, the user may give a mathematical description of the statistical model with \code{setModel} which prepares a \code{yuima.model} object by filling the appropriate slots. If the aim is the simulation of the solution of the stochastic  differential equation specified in the \code{yuima.model} object then, using the method \code{simulate}, it is possible to obtain one trajectory of the process. As an output, a \code{yuima} object (by ``\code{yuima} object'' we mean a, possibly incomplete, object of class \code{yuima}) is created which contains the original model specified in the \code{yuima.model} object in the slot named \code{model} and two additional slots named \code{data}, for the simulated data, and \code{sampling} which contains the description of the simulation scheme used as well as other informations. The details of \code{simulate} will be explained in Section \ref{sec:simul} along with the use of method \code{setSampling} which allows to specify the type of sampling scheme to be used by the \code{simulate} method.
But \code{yuima} object may contain the slot \code{data} non only as the outcome of \code{simulate} but also because the user decides to analyse its own data. In this case the method \code{setData} is used to transform most types  of \proglang{R} time series object into a a proper \code{yuima.data} object.
When the slots \code{data} and \code{model} are available, many other methods can be used to perform statistical analysis on these SDE's (stochastic differential equation) models. These methods will be discussed in Section \ref{sec5}.
Further, functionals of stochastic differential equations can be defined using the \code{setFunctional} method and evaluated using asymptotic expansion methods as explained in Section \ref{sec4}. The \code{setFunctional} method creates a \code{yuima.functional} object which is included along with a \code{yuima.model}  into a \code{yuima} object in order to be used for the evaluation of its asymptotic expansion.

\subsection{The \code{yuima.model} Class}\label{sec:model}
At present, in \pkg{yuima}  three main classes of stochastic differential equations  can be easily specified.  Here we present  a brief
overview of these models as they will be described in detail in Section \ref{sec3}, but this allows to introduce an overall view of the slots of the \code{yuima.model} class.
In \pkg{yuima} one can describe three main families of stochastic processes at present. These models can be one or multidimensional and eventually described as parametric models. Let $X_0=x_0$ be the initial value of the process, then, the main classes are as follows:
\begin{itemize}
\item diffusion models described as
$$  \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t, \quad X_0=x_0$$
 where $W_t$ is a standard Brownian motion;
\item SDE's driven by fractional Gaussian noise, with $H$ the Hurst parameter, described as
$$ \de X_t=a(t,X_t,\theta)dt + b(t,X_t,\theta)\de W_t^{H}, \quad X_0=x_0;$$ 
\item diffusion process with jumps and  L\'evy processes solution to
$$
\begin{aligned}
\de X_t = & \,\,\, a(t,X_t,\theta)\de t + b(t,X_t,\theta)\de W_t + \int\limits_{|z|>1}\!\!\! c(X_{t-},z)\mu(\de t,\de z) \\
&{}+\!\! \int\limits_{0<|z|\le 1}\!\!\!  c(X_{t-},z)\{\mu(\de t,\de z) - \nu(\de z)\de t\}
\end{aligned}, \quad X_0=x_0;
$$
\end{itemize}
 The functions $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ may have a different number of arguments. For example, if the model is homogeneous in time and drift and diffusion coefficients are entirely specified, then we will use the notaion $a(x)$ and $b(x)$ and describe the diffusion model simply as $\displaystyle \de X_t=a(X_t)dt + b(X_t)\de W_t$. And so forth. Detailed hypotheses and regularity conditions on the coefficients $a(\cdot)$, $b(\cdot)$ and $c(\cdot)$ for each class of models will be given in the next sections. Nevertheless, it is important to remark that these notations only matter to the mathematical description of the model while each coefficient is passed to \pkg{yuima} methods as \proglang{R} mathematical expressions. This means that, for example, $a(t, X_t, \theta) = t\cdot \sqrt{\theta X_t}$ will be passed as \code{t*sqrt(x*theta)}, therefore, the order of the arguments is not relevant to \proglang{R} as well as its mathematical description as long as it is consistent through each specific section. Further, \pkg{yuima} is able to accept any user-specified notation for the state variable $x$ (for $X_t$) and the time variable $t$ and the remaining term of an \proglang{R} expression will be interpreted as parameter as will be explained in Section \ref{sec:diff}.
We are now able to give an overview of the main slots of the most important class of the \pkg{yuima} package.

@@

To see inside its structure, one can use the \proglang{R} command \code{str} on a \code{yuima} object.

@@

As seen in the above, the parameter space is accurately described internally in a \code{yuima} object because in inference for stochastic differential equations, estimators of different parameters have different properties. Usually, the rate of convergence for estimators in the diffusion coefficient are similar to the ones in the i.i.d. (independent and identically distributed) sampling while estimators of parameters in the drift coefficient are slower or, in some cases, not even consistent. The \pkg{yuima} package always tries to implement the best statistical inference for the given model under the observed sampling scheme.

@@

\subsection{One Dimensional Diffusion Processes}\label{sec:diff}
Assume that we want to describe the following stochastic differential equation
$$\de X_t = -3 X_t \de t + \frac{1}{1+X_t^2}\de W_t.$$
In the above $a(x) = -3 x$ and $b(x) = \frac{1}{1+x^2}$ according to the notation of previous section and $W_t$ is a standard Wiener process.
This can be described in \pkg{yuima} by specifying the drift and diffusion coefficients as plain
\proglang{R} expressions passed as strings
<<print=FALSE,echo=FALSE,results=hide>>=
library(yuima)
@ 
<<echo=TRUE, print=FALSE,results=hide>>=
mod1 <- setModel(drift = "-3*x", diffusion = "1/(1+x^2)")
@ 
By default, \pkg{yuima} assumes that the state variable is \code{x} and the time variable is \code{t} and the solve variable is again \code{x}. Notice that the left hand-side of the equation is implicit, this is why \code{yuima.model} has the slot \code{solve.variable}.
The user should not be worried about the warning raised by \pkg{yuima} at this stage, as this is just to inform her on the implicit assumption on the solution variable.
At this point, the package fills the proper slots of the \code{yuima} object
<<>>=
str(mod1)
@
